---
title: "Social Sciences Intro to Statistics"
subtitle: "Pset 4: Due MONTH, DATE, YEAR at 11:59pm"
author: "Belle Lee"
date: "06.15.24"
format: pdf
editor: source
---

```{r, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", highlight = TRUE, warning = FALSE, message = FALSE)
  #comment = "#>" makes it so results from a code chunk start with "#>"; default is "##"
```

# Overview 

Welcome to your fourth and last pset of the course. This problem set is intended to give you some practice becoming familiar with regression. In this problem set, we are asking you to: create an R project, render your file, load and investigate an R data frame that is stored on the web, and apply some basic functions to better understand distributions.

- Note: Change the values of the YAML header above to your name and the date. 

## Question 1: Creating an R project

**Create an R project**  

- Create a folder where you want to save files associated with problem set 4. Let's call that folder "problemset4", but you can name it whatever you want.
  - For instance, it could be SSS >> problem_sets >> problemset3.
- In RStudio, click on "File" >> "New Project" >> "Existing Directory" >> "Browse".
- Browse to find and select your problem set 4 folder.
- Click on "Create Project".
  - An R project file has the extension ".Rproj".
  - The name of the file should be "problemset4.Rproj", or whatever you named the folder.

Save this problemset4.qmd file anywhere in the folder named problemset4.

- At the top of this .qmd file, type in your first and last name in the appropriate place in the YAML header (e.g. "Belle Lee").
- in the date field of the YAML header, insert the date within quotations (any date format is fine).
- Now click the "Render" button near the top of your RStudio window (icon with blue arrow sign) or drop down "File" and select "Render Document".
  - Alternatively you can use the shortcut: __Cmd/Ctrl + Shift + k__.

  - *Note*: One goal of this assignment is to make sure you are able to render without running into errors.
  
## Question 2: Bivariate Regression

1. Describe what it means if the relationship between X and Y is negative? Provide an example.

**ANSWER:**


2. What is a residual? 

**ANSWER:**



3. What is covariance? What is correlation?  

**ANSWER:**


4. Would you determine the following values of correlation as weak, moderate, or strong?

a. 0.15 
b. -0.8
c. 0.77
d. 0.46

**ANSWER:**



## Question 3: SER

1. What is SER? What does a high SER mean? Should we be concerned?

**ANSWER:**



2. Using the syntax for "Residual Standard Error", what is the syntax to reference SER in r? 

**ANSWER:**



## Question 4: Regression
1. Load the ipeds data below from the course website.
```{r}
library(tidyverse)
library(ggplot2)
library(labelled)
library(patchwork)

# Load ipeds dataset from course website
load(url('https://raw.githubusercontent.com/bcl96/Social-Sciences-Stats/main/data/ipeds/output_data/panel_data.RData'))

# Create ipeds data frame with fewer variables/observations
df_ipeds_pop <- panel_data %>%
  # keep data from fall 2022
  filter(year == 2022) %>%
  # which universities to keep:
    # 2015 carnegie classification: keep research universities (15,16,17) and master's universities (18,19,20)
  filter(c15basic %in% c(15,16,17,18,19,20)) %>%
  # which variables to keep
  select(instnm,unitid,opeid6,opeid,control,c15basic,stabbr,city,zip,locale,obereg, # basic institutional characteristics
         tuition6,fee6,tuition7,fee7, # avg tuition and fees for full-time grad, in-state and out-of-state
         isprof3,ispfee3,osprof3,ospfee3, # avg tuition and fees for MD, in-state and out-of-state
         isprof9,ispfee9,osprof9,ospfee9, # avg tuition and fees for Law, in-state and out-of-state
         chg4ay3,chg7ay3,chg8ay3) %>% # [undergraduate] books+supplies; off-campus (not with family) room and board; off-campus (not with family) other expenses
  # rename variables; syntax <new_name> = <old_name>
  rename(region = obereg, # revion
         tuit_grad_res = tuition6, fee_grad_res = fee6, tuit_grad_nres = tuition7, fee_grad_nres = fee7, # grad
         tuit_md_res = isprof3, fee_md_res = ispfee3, tuit_md_nres = osprof3, fee_md_nres = ospfee3, # md
         tuit_law_res = isprof9, fee_law_res = ispfee9, tuit_law_nres = osprof9, fee_law_nres = ospfee9, # law
         books_supplies = chg4ay3, roomboard_off = chg7ay3, oth_expense_off = chg8ay3) %>% # [undergraduate] expenses
  # create measures of tuition+fees
  mutate(
    tuitfee_grad_res = tuit_grad_res + fee_grad_res, # graduate, state resident
    tuitfee_grad_nres = tuit_grad_nres + fee_grad_nres, # graduate, non-resident
    tuitfee_md_res = tuit_md_res + fee_md_res, # MD, state resident
    tuitfee_md_nres = tuit_md_nres + fee_md_nres, # MD, non-resident
    tuitfee_law_res = tuit_law_res + fee_law_res, # Law, state resident
    tuitfee_law_nres = tuit_law_nres + fee_law_nres) %>% # Law, non-resident  
  # create measures of cost-of-attendance (COA) as the sum of tuition, fees, book, living expenses
  mutate(
    coa_grad_res = tuit_grad_res + fee_grad_res + books_supplies + roomboard_off + oth_expense_off, # graduate, state resident
    coa_grad_nres = tuit_grad_nres + fee_grad_nres + books_supplies + roomboard_off + oth_expense_off, # graduate, non-resident
    coa_md_res = tuit_md_res + fee_md_res + books_supplies + roomboard_off + oth_expense_off, # MD, state resident
    coa_md_nres = tuit_md_nres + fee_md_nres + books_supplies + roomboard_off + oth_expense_off, # MD, non-resident
    coa_law_res = tuit_law_res + fee_law_res + books_supplies + roomboard_off + oth_expense_off, # Law, state resident
    coa_law_nres = tuit_law_nres + fee_law_nres + books_supplies + roomboard_off + oth_expense_off) %>% # Law, non-resident    
  # keep only observations that have non-missing values for the variable coa_grad_res
    # this does cause us to lose some interesting universities, but doing this will eliminate some needless complications with respect to learning core concepts about statistical inference
  filter(!is.na(coa_grad_res))

# Add variable labels to the tuit+fees variables and coa variables
  # tuition + fees variables
    var_label(df_ipeds_pop[['tuitfee_grad_res']]) <- 'graduate, full-time, resident; avg tuition + required fees'
    var_label(df_ipeds_pop[['tuitfee_grad_nres']]) <- 'graduate, full-time, non-resident; avg tuition + required fees'
    var_label(df_ipeds_pop[['tuitfee_md_res']]) <- 'MD, full-time, state resident; avg tuition + required fees'
    var_label(df_ipeds_pop[['tuitfee_md_nres']]) <- 'MD, full-time, non-resident; avg tuition + required fees'
    var_label(df_ipeds_pop[['tuitfee_law_res']]) <- 'Law, full-time, state resident; avg tuition + required fees'
    var_label(df_ipeds_pop[['tuitfee_law_nres']]) <- 'Law, full-time, non-resident; avg tuition + required fees'
    
  # COA variables
    var_label(df_ipeds_pop[['coa_grad_res']]) <- 'graduate, full-time, state resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'
    var_label(df_ipeds_pop[['coa_grad_nres']]) <- 'graduate, full-time, non-resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'
    var_label(df_ipeds_pop[['coa_md_res']]) <- 'MD, full-time, state resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'
    var_label(df_ipeds_pop[['coa_md_nres']]) <- 'MD, full-time, non-resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'
    var_label(df_ipeds_pop[['coa_law_res']]) <- 'Law, full-time, state resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'
    var_label(df_ipeds_pop[['coa_law_nres']]) <- 'Law, full-time, non-resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'

df_ipeds_pop %>% glimpse()

##########
########## Create data frame of generated variables, with each variable meant to represent the entire population
##########


num_obs <- 10000

# Generate normal distribution w/ custom mean and sd
set.seed(124)
norm_dist <- rnorm(n = num_obs, mean = 50, sd = 5)

# Generate right-skewed distribution
set.seed(124)
rskew_dist <- rbeta(n = num_obs, shape1 = 2, shape2 = 5)

# Generate left-skewed distribution
set.seed(124)
lskew_dist <- rbeta(n = num_obs, shape1 = 5, shape2 = 2)

# Generate standard normal distribution (default is mean = 0 and sd = 1)
set.seed(124)
stdnorm_dist <- rnorm(n = num_obs, mean = 0, sd = 1)  # equivalent to rnorm(10)

# Create dataframe
df_generated_pop <- data.frame(norm_dist, rskew_dist, lskew_dist, stdnorm_dist)

# drop individual objects associated with each variable
rm(norm_dist,rskew_dist,lskew_dist,stdnorm_dist)
rm(num_obs)


##########
########## Create sample versions of generated population data frame and IPEDS population data frame
##########

# create sample version of our generated data
  set.seed(124) # set seed so that everyone ends up with the same random sample
  
  df_generated_sample <- df_generated_pop %>% sample_n(size = 200)
  df_generated_sample %>% glimpse()


# create sample version of our ipeds data

  set.seed(124) # set seed so that everyone ends up with the same random sample
  
  df_ipeds_sample <- df_ipeds_pop %>% sample_n(size = 200) 
  
  # compare mean of coa_grad_res between population and sample
  mean(df_ipeds_pop$coa_grad_res, na.rm = TRUE)
  mean(df_ipeds_sample$coa_grad_res, na.rm = TRUE)

# filtered data frame for categorical based on institution control
df_ipeds_filtered <- df_ipeds_pop %>% filter(control %in% c(1, 2))

# these two vars in the model
df_ipeds_filtered %>% select(control,tuit_grad_res) %>% glimpse()
```

2. Let's investigate the out-of-state cost for full-time graduate students in `df_ipeds_filtered`. To investigate the relationship between institutional control and out-of-state cost for full-time graduate student, which is $X$ and which is $Y$?


**ANSWER:**



3. Show what the average cost, separated for public vs. private would be
```{r}

```

4. Run the regression model. Interpret your findings.
```{r}

```
**ANSWER:**



## Question 4: Confidence Interval

1. What is a confidence interval? Give an example of how you would phrase findings that fall within a 95% confidence interval.

**ANSWER:**



2. Calculate the both the 95% and 99% confidence interval for the relationship between institutional control and out-of-state cost for full-time graduate students in `df_ipeds_filtered`. Interpret your findings.

```{r}


```
**ANSWER:**




# Render to pdf and submit problem set 

**Render to pdf** by clicking the "Render" button near the top of your RStudio window (icon with blue arrow) or drop down "File" and select "Render to PDF"

- Go to the [class website] (Need to fill in classwebsite) and under the "Readings & Assignments" >> "Week 9" tab, click on the "Problem set 4 submission link"  
- Submit both .qmd and pdf files  
- Use this naming convention "lastname_firstname_ps#" for your .qmd and pdf files (e.g. lee_belle_ps4.qmd & lee_belle_ps4.pdf)  
