---
title: "In-class exercise: Think of research questions and variables of interest using ELS:2002 data"
urlcolor: blue
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true 
    number_sections: true
    highlight: tango 
    df_print: tibble #options: default, tibble, paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# remove scientific notation
options(scipen=999)

options(tibble.width = Inf, width = 10000) # Code necessary to enable scroll right for printing of data frames
```

<!-- Code to enable scroll right for printing of data frames -->
<style>
pre code, pre, code {
  white-space: pre !important;
  overflow-x: auto !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
</style>

# Instructions  

- Run the below code, which loads libraries and data
- Read the section "ELS:2002 data", which provides and introduction to ELS:2002 and also presents some example causal research questions using ELS:2002 data (run the code chunks on your own)
- In the section "Think of research questions and variable", your task is to:
  - write your notes/ideas about dependent variables of interest, independent variables of interest, research questions of interest, populations of interest in the google doc for your group


## Load libraries and data

Please run the code in the following chunk, which does the following:

- Loads libraries (install if necessary)
- Runs *R* script that creates data frames from ELS:2002 data


```{r, include= FALSE}
#-----------------------------------------------------------------------------
# Load libraries
#-----------------------------------------------------------------------------

  # install packages if you haven't already
    #install.packages('tidyverse')
    #install.packages('labelled')
    #install.packages('haven')

  library(tidyverse)
  library(labelled)
  library(haven)


#-----------------------------------------------------------------------------
# LOAD ELS DATA
#-----------------------------------------------------------------------------

# RUN SCRIPT THAT CREATES STUDENT-LEVEL DATA FRAME CONTAINING ALL VARIABLES AND CREATES DATA FRAME WITH A SUBSET OF VARIABLES

  #NOTE: this script will take 30 seconds to a minute to run because loading a dataset w/ about 16,000 observations and 4,000 variables from a website

  source(file = url('https://github.com/anyone-can-cook/educ152/raw/main/scripts/els/read_els_by_pets.R'))
    #source(file = file.path('.','..','..','scripts','els','read_els_by_pets.R'))

# Create a dataframe df_els_stu_fac that has categorical variables as factor class variables rather than labelled class variables
  df_els_stu_fac <- as_factor(df_els_stu, only_labelled = TRUE)
  # convert continuous variables we know we want numeric back to numeric
  for (v in c('bytxmstd','bytxrstd','f1txmstd','f3stloanamt','f3stloanpay','f3ern2011','f3tzrectrans','f3tzreqtrans','f3tzschtotal')) {
    df_els_stu_fac[[v]] <- df_els_stu[[v]]  
  }
```

# ELS:2002 data

## Intro to ELS:2002 data

<br>

The Educational Longitudinal Study (ELS) of 2002, referred to as ELS:2002

- Nationally representative, longitudinal survey of US 10th graders in 2002, followed them from 2002 to 2013
- Created by the National Center for Education Statistics, a branch of the U.S. Department of Education
- [LINK](https://nces.ed.gov/surveys/els2002/) to ELS:2002 website
- [LINK](https://nces.ed.gov/OnlineCodebook) to the NCES "Online Codebook"  website, which contains "public-use" versions of ELS datasets and detailed information about all variables
  - variables that could potentially be used to identify ELS survey respondents are "suppressed" from public-use datasets
  - these variables are present in "restricted" versions of ELS datasets

  
<br>  

ELS datasets we created

- student-level data frame `df_els_stu_all` has all variables (about 4,000) and all observations (about 16,200 observations); one observation for each student
- student-level data frame `df_els_stu` has a subset of variables and a subset of observations
  - sample is respondents who filled completed all waves of the survey; who attended postsecondary education at some point by 2012; and some other minor sample criteria around observations not being missing for certain variables
- for both `df_els_stu_all` and `df_els_stu`, most categorical variables are `labelled` class variables
- student-level data frame `df_els_stu_fac` is exact same as `df_els_stu` except categorical variables are `factor` class; categorical variables must be factor class when running regression models
- In the coming weeks, we want your help in identifying variables of interest and populations/samples of interest!
<br>  

Investigate `df_els_stu` (output not shown)
```{r, include = FALSE}
# variable names
df_els_stu %>% glimpse()

# variable labels
df_els_stu %>% var_label()
```

<br>  

Investigate `df_els_stu_fac`
```{r}
# variable names
df_els_stu_fac %>% glimpse()

# variable labels
df_els_stu_fac %>% var_label()
```


<br>  



## Example ELS:2002 research questions

For the multivariate regression lecture, we will focus on using multivariate regression to answer *causal* research questions that follow the form "what is the effect of $X$ on $Y$?"

<br>

Example research questions using ELS:2002 data


1. [observational design, categorical $X$] What is the effect of having an internship or co-op job ($X$) (in 2005-06) on 2011 earnings ($Y$)? [for students who were enrolled in postsecondary education in 2005-06]
    - $Y$ is `f3ern2011`
    - $X$ is `f2intern0506` (this would be sophomore year for students who graduated HS in spring 2004 and immediately entered postsecondary education)
      - `f2intern0506==0`: no co-op/internship
    - sample: `f2enroll0506=='yes'`
    
```{r}
df_els_stu_fac %>% filter(f2enroll0506=='yes') %>% 
  group_by(f2intern0506) %>% summarize(
  mean_earn = mean(f3ern2011, na.rm = TRUE),
  n_nonmissing = sum(!is.na(f3ern2011))
)
```
    

<br><br>

2. [observational design, categorical $X$] What is the effect of field of study ($X$) on earnings ($Y$)? [for respondents not currently enrolled in postsecondary education]
    - answer the RQ separately for: certificate degrees; associate's degrees; for bachelor's degrees?
    - $Y$ is `f3ern2011`
    - for certificate degrees: $X$ is `f3tzcrt1cip2`
    - for associate's degrees: $X$ is `f3tzasc1cip2`
    - for bachelor's degrees: $X$ is `f3tzbch1cip2 `

```{r}
# certificate degrees
df_els_stu_fac %>% filter(!(f3tzcrt1cip2 %in% c('Missing','Item legitimate skip/NA')), f3a01d=='No') %>% 
  group_by(f3tzcrt1cip2) %>% summarize(
  mean_earn = mean(f3ern2011, na.rm = TRUE),
  n_nonmissing = sum(!is.na(f3ern2011))
)

# associates degreees
df_els_stu_fac %>% filter(!(f3tzasc1cip2 %in% c('Missing','Item legitimate skip/NA')), f3a01d=='No') %>% 
  group_by(f3tzasc1cip2) %>% summarize(
  mean_earn = mean(f3ern2011, na.rm = TRUE),
  n_nonmissing = sum(!is.na(f3ern2011))
)

# bachelor's degreees
df_els_stu_fac %>% filter(!(f3tzbch1cip2 %in% c('Missing','Item legitimate skip/NA')), f3a01d=='No') %>% 
  group_by(f3tzbch1cip2) %>% summarize(
  mean_earn = mean(f3ern2011, na.rm = TRUE),
  n_nonmissing = sum(!is.na(f3ern2011))
) %>% print(n=30)
```
<br><br>

3. [observational design, continuous $X$] What is the effect of amount borrowed in student loans ($X$) on earnings ($Y$) for respondents not currently enrolled in postsecondary education
    - $X$ var is `f3totloan`; $Y$ var is `f3ern2011`
    - sample: `f3a01d=='No'`


```{r}
df_els_stu_fac %>% filter(f3a01d=="No") %>%  ggplot(aes(x=f3totloan, y=f3ern2011)) + geom_point() + stat_smooth(method = 'lm')

summary(lm(formula = f3ern2011 ~ f3totloan, data = df_els_stu %>% filter(f3a01d==0)))
```

<br>
<br>
Why are we using earnings as the dependent variable?

- Ordinary Least Squares (OLS) multivariate regression assumes continuous dependent variable; regression models with categorical dependent variables are a bit more difficult to interpret
- But in later weeks, we may introduce the "linear probability model," which will enable us to model dichotomous (two-category) dependent variables (e.g., 0/1 did student graduate)

# Think of variables and research questions of interest

Your task is to:

- think of potential dependent variables of interest and potential independent variables of interest
  - some of these variables may already be in the data frame `els_stu_fac`; some may be in the larger data frame `els_stu_all`, and I can add them to `els_stu_fac` in subsequent weeks
  - dependent variables can be continous or dichotomous (two categories)
  - independent variables should be variables that could be (in theory) randomly assigned in a random assignment experiment
- think of potential causal research questions of interest that follow the form "what is the effect of X on Y"
  - you may think about what sample/population of students this research questions pertains to
- write your notes/ideas about dependent variables of interest, independent variables of interest, research questions of interest, populations of interest in the google doc for your group;
  - doesn't have to be polished! But I'll use your ideas when thinking about which variables to add to `els_stu_fac`, and when thinking about questions for the take-home final exam (I want research questions that speak to your interests)
  
<br>

Can use the `lookfor()` function from the `labelled` package to query which variables and variable labels contain certain words

- syntax (with defaults):
  - `look_for(data, ..., labels = TRUE, ignore.case = TRUE, details = TRUE)`
  - where argument `...` contains "strings" (e.g., words) you are looking for, enclosed in quotation marks
- example syntax, looking for variable names and variable labels that contain the word "internship"
  - `lookfor(data=df_els_stu_all, 'internship',labels = TRUE, ignore.case = TRUE, details = TRUE)`

<br>

useful to use `lookfor()` to query the dataset `df_els_stu_all` because this dataset has so many variables (more than 4,000)
```{r}

# query data frame df_els_stu_all
  lookfor(data=df_els_stu_all, 'internship',labels = TRUE, ignore.case = TRUE, details = TRUE)


# query data frame df_els_stu
  lookfor(data=df_els_stu, 'debt',labels = TRUE, ignore.case = TRUE, details = TRUE)
  lookfor(data=df_els_stu, 'loan',labels = TRUE, ignore.case = TRUE, details = TRUE)
```

<br>

Investigate `df_els_stu_fac` (output not shown)
```{r, include = FALSE}
# variable names
df_els_stu_fac %>% glimpse()

# variable labels
df_els_stu_fac %>% var_label()
```