---
title: "In-class exercise: Objects, Tidyverse, & Pipes"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions  
- Take a few minutes to review section 3, Tidyverse & Pipes, from the R walkthrough document [here](https://anyone-can-cook.github.io/educ152/assets/resources/educ152_walkthrough.html#3_Tidyverse_and_Pipes)  

### Task  

__Install/load packages and read in data__  

- Run the code chunks below to install and load `Tidyverse`.  
You will be reading in a subset of the college scorecard data from lecture that includes the following 8 universities:  
    - UC-Berkeley = 001312
    - UCLA = 001315
    - USC = 001328
    - Stanford = 001305
    - Columbia = 002707
    - Columbia, Teacher's College = 003979
    - NYU = 002785
    - Harvard = 002155.
    
Install Tidyverse (if you haven't already) and load the library.
```{r message=FALSE}
#install.packages("tidyverse") #uncomment this line out
library(tidyverse)
```

Read in a subset of the college scorecard data
```{r}
#Read in data set (data frame)
load(url("https://github.com/anyone-can-cook/educ152/raw/main/data/college_scorecard/output_data/scorecard_edu_small.RData"))

#print first 6 observations or rows
scorecard_edu_sm
```

__Pipes__  

- Now that we have `Tidyverse` installed and loaded, we can begin to use `pipes`. 
- Run the code chunk below using the `names()` to print the variable names in the `scorecard_edu_sm` dataframe.  
```{r}
names(scorecard_edu_sm)
```

- Say we wanted to view only the institution names. Run the code chunk below using the `select()` function to select the column or variable `instnm`.
```{r}
scorecard_edu_sm %>% select(instnm)
```

- What if I only wanted to see universities in California? Run the code chunk below using the `filter()` function to filter for rows or observations of universities in California.
```{r}
scorecard_edu_sm %>% select(instnm, stabbr) %>% filter(stabbr == "CA")
```

- Say I wanted to create an object of universities in CA. I would have to assign it using the assignment operator `<-` or `=`. The `cali_schools` is an object I created below.  
```{r}
cali_schools <- scorecard_edu_sm %>% filter(stabbr == "CA") %>% select(instnm, stabbr) %>% unique() #using the unique() function to only grab unique rows/observations
```

- It is considered a `list` object.
```{r}
typeof(cali_schools)
class(cali_schools)
```

### Now it is your turn:

__Question 1:__ Using the dataframe `scorecard_edu_sm`, select the variables/columns `instnm`, `stabbr`, and `control`. Don't forget to use a pipe `%>%`


```{r}

```

<br>

__Question 2:__ Using the same code from above, filter for universities in the state `stabbr` of "NY". Don't forget to use a pipes `%>%` in your code.


```{r}

```
 
<br>

__Question 3:__ Now say we want to save our code from above as an object. Create an object called `ny_schools`. Don't forget to use a pipes `%>%` in your code. _Hint: you maybe want to use unique() at the end of your code_


```{r}

```

<br>


