---
title: "In-class exercise: Distributions and Sampling distributions"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions  

- We are going to work through some concepts discussed in the previous week's lecture (e.g. sampling distribution, central limit theorem, working with user-written functions). 

### Task 

Please run the code in the following chunk, which does the following:

- Loads libraries
- Loads and creates IPEDS data frame  (population)
- Creates data frame of generated variables (population)
- Creates sample versions of the IPEDS and gnerated data frames

_Note: code chunk omitted from html document using include = FALSE_
```{r, include= FALSE}
#-----------------------------------------------------------------------------
# Load libraries
#-----------------------------------------------------------------------------

  #install.packages('tidyverse') # if you haven't installed already
  #install.packages('labelled') # if you haven't installed already
  #install.packages('patchwork') # if you haven't installed already

library(tidyverse) # load tidyverse package
library(labelled) # load labelled package package
library(patchwork)

#-----------------------------------------------------------------------------
# Load ipeds dataset from course website url
#-----------------------------------------------------------------------------

load(file = url('https://github.com/anyone-can-cook/educ152/raw/main/data/ipeds/output_data/panel_data.RData'))

# Create ipeds data frame with fewer variables/observations
df_ipeds_pop <- panel_data %>%
  # keep data from fall 2019
  filter(year == 2019) %>%
  # which universities to keep:
    # 2015 carnegie classification: keep research universities (15,16,17) and master's universities (18,19,20)
  filter(c15basic %in% c(15,16,17,18,19,20)) %>%
  # which variables to keep
  select(instnm,unitid,opeid6,opeid,control,c15basic,stabbr,city,zip,locale,obereg, # basic institutional characteristics
         tuition6,fee6,tuition7,fee7, # avg tuition and fees for full-time grad, in-state and out-of-state
         isprof3,ispfee3,osprof3,ospfee3, # avg tuition and fees for MD, in-state and out-of-state
         isprof9,ispfee9,osprof9,ospfee9, # avg tuition and fees for Law, in-state and out-of-state
         chg4ay3,chg7ay3,chg8ay3) %>% # [undergraduate] books+supplies; off-campus (not with family) room and board; off-campus (not with family) other expenses
  # rename variables; syntax <new_name> = <old_name>
  rename(region = obereg, # revion
         tuit_grad_res = tuition6, fee_grad_res = fee6, tuit_grad_nres = tuition7, fee_grad_nres = fee7, # grad
         tuit_md_res = isprof3, fee_md_res = ispfee3, tuit_md_nres = osprof3, fee_md_nres = ospfee3, # md
         tuit_law_res = isprof9, fee_law_res = ispfee9, tuit_law_nres = osprof9, fee_law_nres = ospfee9, # law
         books_supplies = chg4ay3, roomboard_off = chg7ay3, oth_expense_off = chg8ay3) %>% # [undergraduate] expenses
  # create measures of tuition+fees
  mutate(
    tuitfee_grad_res = tuit_grad_res + fee_grad_res, # graduate, state resident
    tuitfee_grad_nres = tuit_grad_nres + fee_grad_nres, # graduate, non-resident
    tuitfee_md_res = tuit_md_res + fee_md_res, # MD, state resident
    tuitfee_md_nres = tuit_md_nres + fee_md_nres, # MD, non-resident
    tuitfee_law_res = tuit_law_res + fee_law_res, # Law, state resident
    tuitfee_law_nres = tuit_law_nres + fee_law_nres) %>% # Law, non-resident  
  # create measures of cost-of-attendance (COA) as the sum of tuition, fees, book, living expenses
  mutate(
    coa_grad_res = tuit_grad_res + fee_grad_res + books_supplies + roomboard_off + oth_expense_off, # graduate, state resident
    coa_grad_nres = tuit_grad_nres + fee_grad_nres + books_supplies + roomboard_off + oth_expense_off, # graduate, non-resident
    coa_md_res = tuit_md_res + fee_md_res + books_supplies + roomboard_off + oth_expense_off, # MD, state resident
    coa_md_nres = tuit_md_nres + fee_md_nres + books_supplies + roomboard_off + oth_expense_off, # MD, non-resident
    coa_law_res = tuit_law_res + fee_law_res + books_supplies + roomboard_off + oth_expense_off, # Law, state resident
    coa_law_nres = tuit_law_nres + fee_law_nres + books_supplies + roomboard_off + oth_expense_off) %>% # Law, non-resident    
  # keep only observations that have non-missing values for cost of off-campus room and board and cost of off-campus other expenses and non-missing values for tuit+fees
  filter(!is.na(roomboard_off) & !is.na(oth_expense_off) & !is.na(tuitfee_grad_nres))

# Add variable labels to the tuit+fees variables and coa variables
  # tuition + fees variables
    var_label(df_ipeds_pop[['tuitfee_grad_res']]) <- 'graduate, full-time, resident; avg tuition + required fees'
    var_label(df_ipeds_pop[['tuitfee_grad_nres']]) <- 'graduate, full-time, non-resident; avg tuition + required fees'
    var_label(df_ipeds_pop[['tuitfee_md_res']]) <- 'MD, full-time, state resident; avg tuition + required fees'
    var_label(df_ipeds_pop[['tuitfee_md_nres']]) <- 'MD, full-time, non-resident; avg tuition + required fees'
    var_label(df_ipeds_pop[['tuitfee_law_res']]) <- 'Law, full-time, state resident; avg tuition + required fees'
    var_label(df_ipeds_pop[['tuitfee_law_nres']]) <- 'Law, full-time, non-resident; avg tuition + required fees'
    
  # COA variables
    var_label(df_ipeds_pop[['coa_grad_res']]) <- 'graduate, full-time, state resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'
    var_label(df_ipeds_pop[['coa_grad_nres']]) <- 'graduate, full-time, non-resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'
    var_label(df_ipeds_pop[['coa_md_res']]) <- 'MD, full-time, state resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'
    var_label(df_ipeds_pop[['coa_md_nres']]) <- 'MD, full-time, non-resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'
    var_label(df_ipeds_pop[['coa_law_res']]) <- 'Law, full-time, state resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'
    var_label(df_ipeds_pop[['coa_law_nres']]) <- 'Law, full-time, non-resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'

df_ipeds_pop %>% glimpse()

#-----------------------------------------------------------------------------
# Create data frame of generated variables, with each variable meant to represent the entire population
#-----------------------------------------------------------------------------

num_obs <- 10000

# Generate normal distribution w/ custom mean and sd
set.seed(124)
norm_dist <- rnorm(n = num_obs, mean = 50, sd = 5)

# Generate right-skewed distribution
set.seed(124)
rskew_dist <- rbeta(n = num_obs, shape1 = 2, shape2 = 5)

# Generate left-skewed distribution
set.seed(124)
lskew_dist <- rbeta(n = num_obs, shape1 = 5, shape2 = 2)

# Generate standard normal distribution (default is mean = 0 and sd = 1)
set.seed(124)
stdnorm_dist <- rnorm(n = num_obs, mean = 0, sd = 1)  # equivalent to rnorm(10)

# Create dataframe
df_generated_pop <- data.frame(norm_dist, rskew_dist, lskew_dist, stdnorm_dist)

# drop individual objects associated with each variable
rm(norm_dist,rskew_dist,lskew_dist,stdnorm_dist)
rm(num_obs)

#-----------------------------------------------------------------------------
# Create sample versions of generated population data frame and IPEDS #-----------------------------------------------------------------------------

# create sample version of our generated data
  set.seed(124) # set seed so that everyone ends up with the same random sample
  
  df_generated_sample <- df_generated_pop %>% sample_n(size = 200)
  df_generated_sample %>% glimpse()


# create sample version of our ipeds data

  set.seed(124) # set seed so that everyone ends up with the same random sample
  
  df_ipeds_sample <- df_ipeds_pop %>% sample_n(size = 200) 


#-----------------------------------------------------------------------------
# Create function to generate plots of variable distributions
#-----------------------------------------------------------------------------

plot_distribution <- function(data_vec, plot_title = '') {
  p <- ggplot(as.data.frame(data_vec), aes(x=data_vec)) + xlab('') +
    ggtitle(plot_title) +
    geom_histogram(aes(y=..density..), alpha=0.4, position='identity') +
    geom_density() +
    geom_vline(aes(xintercept=mean(data_vec, na.rm = T), color='mean'),
               linetype='dotted', size=0.8, alpha=0.8) +
    geom_vline(aes(xintercept=median(data_vec, na.rm = T), color='median'),
               linetype='dotted', size=0.8, alpha=0.8) +
    scale_color_manual(name = 'Statistics',
                       labels = c(paste('Mean:', round(mean(data_vec, na.rm = T), 2),
                                        '\nStd Dev:', round(sd(data_vec, na.rm = T), 2)), 
                                paste('Median:', round(median(data_vec, na.rm = T), 2))),
                       values = c(median='blue', mean='red')) +
    theme(plot.title = element_text(size=10, face='bold', hjust = 0.5),
          legend.title = element_text(size=9, face='bold'),
          legend.text = element_text(size=8))
  
  p
}


#-----------------------------------------------------------------------------
# Write function to get the sampling distribution from a variable (defaults equal 500 samples of size 200)
#-----------------------------------------------------------------------------

get_sampling_distribution <- function(data_vec, num_samples = 500, sample_size = 200) {
  sample_means <- vector(mode = 'numeric', num_samples)

  for (i in 1:length(sample_means)) {
    samp <- sample(data_vec, sample_size)
    sample_means[[i]] <- mean(samp, na.rm = T)
  }

  sample_means
}

#-----------------------------------------------------------------------------
# Write Function to generate sampling distribution (with t-test value) assuming null hypothesis is correct
#-----------------------------------------------------------------------------


plot_t_distribution <- function(data_vec, mu, alpha = 0.05, alternative = 'two.sided', plot_title = '') {
  
  data_vec <- na.omit(data_vec)
  
  # Calculate t-statistics
  sample_size <- length(data_vec)
  deg_freedom <- sample_size - 1
  xbar <- mean(data_vec)
  s <- sd(data_vec)
  
  std_err <- s / sqrt(sample_size)
  t <- (xbar - mu) / std_err
  
  # Calculate critical value and p-value
  if (alternative == 'less') {  # left-tailed
    cv_lower <- qt(p = alpha, df = deg_freedom, lower.tail = T)
    cv_legend <- round(cv_lower, 2)
    cv_legend2 <- round(cv_lower * std_err + mu, 2)
    pval <- pt(q = t, df = deg_freedom, lower.tail = T)
  } else if (alternative == 'greater') {  # right-tailed
    cv_upper <- qt(p = alpha, df = deg_freedom, lower.tail = F)
    cv_legend <- round(cv_upper, 2)
    cv_legend2 <- round(cv_upper * std_err + mu, 2)
    pval <- pt(q = t, df = deg_freedom, lower.tail = F)
  } else {  # two-tailed
    cv_lower <- qt(p = alpha / 2, df = deg_freedom, lower.tail = T)
    cv_upper <- qt(p = alpha / 2, df = deg_freedom, lower.tail = F)
    cv_legend <- str_c('\u00B1', round(cv_upper, 2))
    cv_legend2 <- str_c(round(cv_lower * std_err + mu, 2), ' & ', round(cv_upper * std_err + mu, 2))
    pval <- 2 * pt(q = t, df = deg_freedom, lower.tail = t < 0)
  }
  
  # Plot t-distribution
  p <- ggplot(data.frame(x = -c(-4, 4)), aes(x)) +
    ggtitle(plot_title) + xlab('') + ylab('') +
    stat_function(fun = dt, args = list(df = deg_freedom))
  
  # Shade rejection region using critical value
  if (alternative != 'greater') {
    p <- p +
      stat_function(fun = dt, args = list(df = deg_freedom),
                    xlim = c(-4, cv_lower),
                    geom = 'area', alpha = 0.3, fill = 'red') +
      geom_vline(aes(xintercept = cv_lower, color = 'cval'),
                 linetype = 'dotted', size = 0.8, alpha = 0.8)
  }
  if (alternative != 'less') {
    p <- p +
      stat_function(fun = dt, args = list(df = deg_freedom),
                    xlim = c(cv_upper, 4),
                    geom = 'area', alpha = 0.3, fill = 'red') +
      geom_vline(aes(xintercept = cv_upper, color = 'cval'),
                 linetype = 'dotted', size = 0.8, alpha = 0.8)
  }
  
  # Label plot
  p <- p +
    geom_vline(aes(xintercept = t, color = 'tstat'),
               linetype = 'dotted', size = 0.8, alpha = 0.8) +
    scale_x_continuous(sec.axis = sec_axis(trans = ~ . * std_err + mu)) +
    scale_color_manual(name = 'Legend',
                       breaks = c('tstat', 'cval'),
                       labels = c(str_c('t-statistics: ', round(t, 2), ' / ', round(xbar, 2),
                                        ' (SE: ', round(std_err, 2), ')',
                                        '\n(p-value: ', round(pval, 3), ')'),
                                  str_c('Critical value: ', cv_legend, ' / ', cv_legend2,
                                        '\n(alpha: ', round(alpha, 2), ')')),
                       values = c(tstat='blue', cval='red')) +
    theme(plot.title = element_text(size = 10, face = 'bold', hjust = 0.5),
          legend.title = element_text(size = 9, face = 'bold'),
          legend.text = element_text(size = 8))

  p
}
```

### Question 1

If we wanted to plot the distribution of the variable `coa_grad_nres` from the data frame `df_ipeds_pop`, we could use the user-written function called `plot_distribution` like so:
```{r warning=FALSE, message=FALSE}
plot_distribution(data_vec = df_ipeds_pop$coa_grad_nres, plot_title = "Distribution of cost of attendance for out-of-state graduate students")
```

- Does the distribution above have a normal, left-skewed, or right-skwed shape? Why?
   - __ANSWER__: 


- What is the standard deviation and interpret this value in words?
  - __ANSWER__:


### Question 2

Run the following code, which does the following:

- Takes 1000 random samples of sample size n=20 from the data frame `df_ipeds_pop`.
- For each random sample, calculates the sample mean of variable `coa_grad_nres`.
- Plots the sampling distribution of the sample mean of variable  `coa_grad_nres`.

```{r warning=FALSE, message=FALSE}
set.seed(345)
get_sampling_distribution(data_vec = df_ipeds_pop$coa_grad_nres, num_samples = 1000, sample_size = 20) %>%
  plot_distribution(plot_title = "Sampling Distribution of the Sample mean")
```

- What does each observation in the above plot represent?
  - __ANSWER__:

  
- Would you describe the shape of the above distribtuion as (approximately) normal, left-skwed, or right-skewed?
  - __ANSWER__:

  
__Your turn__  

Run the following in the code chunk below:  

- Take 1000 random samples of sample size n=200 from the data frame `df_ipeds_pop`.
- For each random sample, calculates the sample mean of variable `coa_grad_nres`.
- Plots the sampling distribution of the sample mean of variable  `coa_grad_nres`.

```{r warning=FALSE, message=FALSE}
set.seed(345)

```

  
- Why is the standard error from this sampling distribution (each sample has sample size n=200) different than the standard error from the sampling distribution from the previous example (each sample has sample size n=20)?
  - __ANSWER__:


### Question 3  
Run the following code, which does the following:

- Plots the population distribution of the variable `coa_grad_nres`
- Plots the distribution of the variable  `coa_grad_nres` from one sample
- Plots the sampling distribution of the sample mean for the variable `coa_grad_nres`
```{r warning=FALSE, message=FALSE}
set.seed(345)
plot_distribution(df_ipeds_pop$coa_grad_nres, plot_title = 'Population distribution') +
  plot_distribution(df_ipeds_sample$coa_grad_nres, plot_title = 'Single sample distribution') +
  plot_distribution(get_sampling_distribution(data_vec = df_ipeds_pop$coa_grad_nres, num_samples = 1000, sample_size = 200),plot_title = "sampling distribution of sample mean of coa_grad_nres") +
  plot_layout(ncol = 1)
```

- State the central limit theorem in your own words and explain why it is important for hypothesis testing
  - __ANSWER__:


<br>
