---
title: "Social Sciences Intro to Statistics"
subtitle: "Week 7.2 Introduction to Bivariate Regression (continued)"
format: pdf
editor: source
---
Week 7: Learning goal -  Demonstrate estimation and prediction of bivariate regression analysis in R 

 
```{r, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", highlight = TRUE, warning = FALSE, message = FALSE)
  #comment = "#>" makes it so results from a code chunk start with "#>"; default is "##"
```

# Introduction

Load packages:
```{r, message=FALSE}
library(tidyverse)
library(ggplot2)

```

Resources used to create this lecture:

# Estimation
__General things we do in regression analysis__

1. __Estimation__ 
    - How do we choose estimates of $\beta_0$ and $\beta_1$ using sample data?

<br>

2. __Prediction__ 
    - What is the predicted value of Y for someone with a particular value of X?
  
<br>
  
3. __Hypothesis testing__ [focus of the rest of the semester]
    - Hypothesis testing and confidence intervals about $\beta_1$
  

<br>

Step 1 of regression is to estimate parameters of the population linear regression model

  - $Y_i = \beta_0 + \beta_1X_i + u_i$

<br>

__Goal of estimation: use sample data to estimate population parameters__:

  - Previous lecture: we used sample data on variable $Y$ to estimate population mean, $\mu_Y$
    - $\bar{Y}$ (sample mean of $Y$) is an estimate of $\mu_Y$
  - This lecture: use sample data to estimate the population intercept, $\beta_0$, and the population regression coefficient, $\beta_1$
    - $\hat{\beta_0}$ is an estimate of $\beta_0$
    - $\hat{\beta_1}$ is an estimate of $\beta_1$






