---
title: "Social Sciences Intro to Statistics"
subtitle: "Week 7.1 Introduction to Bivariate Regression"
format: pdf
editor: source
---
Week 7: Learning goal - Demonstrate estimation and prediction of bivariate regression analysis in R.

 
```{r, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", highlight = TRUE, warning = FALSE, message = FALSE)
  #comment = "#>" makes it so results from a code chunk start with "#>"; default is "##"
```

# Introduction

Load packages:
```{r, message=FALSE}
library(tidyverse)
library(ggplot2)
best_netflix <- read_csv("https://raw.githubusercontent.com/bcl96/Social-Sciences-Stats/main/data/netflix_data/Best%20Shows%20Netflix.csv")
```

Resources used to create this lecture:

# Bivariate Regression
“Bivariate regression” refers to regression models with two variables, a $Y$ variable (“dependent variable” or “outcome”) and a single $X$ variable (“independent variable”).

"Multivariate regression" refers to regression models with a $Y$ variable and two or more $X$ variables

This lecture – which we will teach over several weeks – teaches the fundamental concepts of bivariate regression. All of these concepts will be similar when we move on to multivariate regression. 

### Scatterplot
__Relationships between two continuous variables__

Postive relationship, negative relationship, and no relationships

<br>

Relationship between X and Y is positive

- when X is "high", Y tend to be "high"
- when X is "low", Y tends to be "low"
- e.g., number of hours (X) studying and GPA (Y)
- e.g., cost of attendance (X) and student debt (Y)

<br>

Relationship between X and Y is negative

- when X is "high", Y tend to be "low"
- when X is "low", Y tends to be "high"
- e.g., number of school absences and GPA

<br>

No relationship between X and Y

- knowing the value of X gives you does not tell you much about the value of Y
- e.g., amount of ice cream consumed and GPA
  (defined as "research" or "master's" universities by the Carnegie Classification that are 
  
we will use the data frame `df_socialwork` (which combines debt/earnings data from College Scorecard and tuition/cost of attendance data from IPEDS) to run regression models of the relationship between measures of tuition/COA (X variable) and debt/earnings (Y variable) for MA programs in social work





<br>



<br>

Ways to investigate this relationship between X and Y:

- Graphically: scatterplots
- Numerically: covariance (less used), correlation


## Scatterplots

Scatterplots will plot individual observations on an X and Y axis

<br>

Draw scatterplot of X (`coa_grad_res`) and Y (`debt_all_stgp_eval_mean`)
```{r, eval=FALSE}
df_socialwork %>%  ggplot(aes(x=coa_grad_res, y=debt_all_stgp_eval_mean)) + geom_point()
```

<br>

Create scatterplot with "prediction" line
```{r, eval=FALSE}
df_socialwork %>%  ggplot(aes(x=coa_grad_res, y=debt_all_stgp_eval_mean)) + geom_point() + stat_smooth(method = 'lm')
```


### Covariance

### Correlation




