---
title: "Social Sciences Intro to Statistics"
subtitle: "Week 4.2 Fundamentals of Inferential Statistics"
format: pdf
editor: source
---
Week 4: Learning goal - Apply understanding of central limit theorem and sampling distributions towards how to evaluate inferential statistics in R. 

 
```{r, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", highlight = TRUE, warning = FALSE, message = FALSE)
  #comment = "#>" makes it so results from a code chunk start with "#>"; default is "##"
```

# Introduction

Load packages:
```{r, message=FALSE}
library(tidyverse)
library(ggplot2)
best_netflix <- read_csv("https://raw.githubusercontent.com/bcl96/Social-Sciences-Stats/main/data/netflix_data/Best%20Shows%20Netflix.csv")
```

Resources used to create this lecture:

# Fundamentals of inferential statistics
Population parameters

Sample estimates 

Using sample estimates to make inferences about population parameters

	-Infer package (basic hypothesis testing)
		-specify() allows you to specify the variable, or relationship between variables, that you’re interested in.
 		-hypothesize() allows you to declare the null hypothesis.
 		-generate() allows you to generate data reflecting the null hypothesis.
-calculate() allows you to calculate a distribution of statistics from the generated data to form the null distribution.




