---
title: "Social Sciences Intro to Statistics"
subtitle: "Final: Due MONTH, DATE, YEAR at 11:59pm"
author: "Belle Lee"
date: "06.15.24"
format: pdf
editor: source
---

```{r, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", highlight = TRUE, warning = FALSE, message = FALSE)
  #comment = "#>" makes it so results from a code chunk start with "#>"; default is "##"
```

# Overview 

You made it to the end! Congratulations on working hard the last ten weeks in this course. You're so close to the finish line. Similar to your previous psets, this final use the datasets we have been using in the course. However, unlike the psets, you will work on the final on your own. 

- Note: Change the values of the YAML header above to your name and the date. 

## Question 1: Creating an R project

**Create an R project**  

- Create a folder where you want to save files associated with the final. Let's call that folder "Final", but you can name it whatever you want.
  - For instance, it could be SSS >> Final.
- In RStudio, click on "File" >> "New Project" >> "Existing Directory" >> "Browse".
- Browse to find and select your Final folder.
- Click on "Create Project".
  - An R project file has the extension ".Rproj".
  - The name of the file should be "final.Rproj", or whatever you named the folder.

Save this final.Rmd file anywhere in the folder named final.

- At the top of this .qmd file, type in your first and last name in the appropriate place in the YAML header (e.g. "Belle Lee").
- in the date field of the YAML header, insert the date within quotations (any date format is fine).
- Now click the "Render" button near the top of your RStudio window (icon with blue arrow sign) or drop down "File" and select "Render Document".
  - Alternatively you can use the shortcut: __Cmd/Ctrl + Shift + k__.
  
## Question 2: Definitions

Define the following terms:
1. Standard deviation: **ANSWER KEY:**

2. p-value: **ANSWER KEY:**

3. Mean: **ANSWER KEY:**

4. alpha-level: **ANSWER KEY:**

5. Standard error: **ANSWER KEY:**

6. What are the measures of central tendency? Provide an explanation of each. 
**ANSWER KEY:**

7. Explain what is the Central Limit Theorem and why is it important? 
**ANSWER KEY:**

8. Covariance: **ANSWER KEY:**

9. Correlation: **ANSWER KEY:**

10. What are the OLS (Ordinary Least Squares) Assumptions? 
**ANSWER KEY:**

## Question 3: Conduct a two group hypothesis test

1. Load the necessary package(s): `tidyverse`, `ggplot2`, `labelled`, `patchwork`

2. Load the `ipeds` dataframe within this code chunk below.
```{r}
# ANSWER KEY
library(tidyverse)
library(ggplot2)
library(labelled)
library(patchwork)

#load netflix data
load(url('https://raw.githubusercontent.com/bcl96/Social-Sciences-Stats/main/data/ipeds/output_data/panel_data.RData'))
```
  
3. Run the following code chunk (similar to what was provided in past lectures and psets to create ipeds data frame with fewer variables.

```{r, echo=FALSE}
# Create ipeds data frame with fewer variables/observations
df_ipeds_pop <- panel_data %>%
  # keep data from fall 2022
  filter(year == 2022) %>%
  # which universities to keep:
    # 2015 carnegie classification: keep research universities (15,16,17) and master's universities (18,19,20)
  filter(c15basic %in% c(15,16,17,18,19,20)) %>%
  # which variables to keep
  select(instnm,unitid,opeid6,opeid,control,c15basic,stabbr,city,zip,locale,obereg, # basic institutional characteristics
         tuition6,fee6,tuition7,fee7, # avg tuition and fees for full-time grad, in-state and out-of-state
         isprof3,ispfee3,osprof3,ospfee3, # avg tuition and fees for MD, in-state and out-of-state
         isprof9,ispfee9,osprof9,ospfee9, # avg tuition and fees for Law, in-state and out-of-state
         chg4ay3,chg7ay3,chg8ay3) %>% # [undergraduate] books+supplies; off-campus (not with family) room and board; off-campus (not with family) other expenses
  # rename variables; syntax <new_name> = <old_name>
  rename(region = obereg, # revion
         tuit_grad_res = tuition6, fee_grad_res = fee6, tuit_grad_nres = tuition7, fee_grad_nres = fee7, # grad
         tuit_md_res = isprof3, fee_md_res = ispfee3, tuit_md_nres = osprof3, fee_md_nres = ospfee3, # md
         tuit_law_res = isprof9, fee_law_res = ispfee9, tuit_law_nres = osprof9, fee_law_nres = ospfee9, # law
         books_supplies = chg4ay3, roomboard_off = chg7ay3, oth_expense_off = chg8ay3) %>% # [undergraduate] expenses
  # create measures of tuition+fees
  mutate(
    tuitfee_grad_res = tuit_grad_res + fee_grad_res, # graduate, state resident
    tuitfee_grad_nres = tuit_grad_nres + fee_grad_nres, # graduate, non-resident
    tuitfee_md_res = tuit_md_res + fee_md_res, # MD, state resident
    tuitfee_md_nres = tuit_md_nres + fee_md_nres, # MD, non-resident
    tuitfee_law_res = tuit_law_res + fee_law_res, # Law, state resident
    tuitfee_law_nres = tuit_law_nres + fee_law_nres) %>% # Law, non-resident  
  # create measures of cost-of-attendance (COA) as the sum of tuition, fees, book, living expenses
  mutate(
    coa_grad_res = tuit_grad_res + fee_grad_res + books_supplies + roomboard_off + oth_expense_off, # graduate, state resident
    coa_grad_nres = tuit_grad_nres + fee_grad_nres + books_supplies + roomboard_off + oth_expense_off, # graduate, non-resident
    coa_md_res = tuit_md_res + fee_md_res + books_supplies + roomboard_off + oth_expense_off, # MD, state resident
    coa_md_nres = tuit_md_nres + fee_md_nres + books_supplies + roomboard_off + oth_expense_off, # MD, non-resident
    coa_law_res = tuit_law_res + fee_law_res + books_supplies + roomboard_off + oth_expense_off, # Law, state resident
    coa_law_nres = tuit_law_nres + fee_law_nres + books_supplies + roomboard_off + oth_expense_off) %>% # Law, non-resident    
  # keep only observations that have non-missing values for the variable coa_grad_res
    # this does cause us to lose some interesting universities, but doing this will eliminate some needless complications with respect to learning core concepts about statistical inference
  filter(!is.na(coa_grad_res))

# Add variable labels to the tuit+fees variables and coa variables
  # tuition + fees variables
    var_label(df_ipeds_pop[['tuitfee_grad_res']]) <- 'graduate, full-time, resident; avg tuition + required fees'
    var_label(df_ipeds_pop[['tuitfee_grad_nres']]) <- 'graduate, full-time, non-resident; avg tuition + required fees'
    var_label(df_ipeds_pop[['tuitfee_md_res']]) <- 'MD, full-time, state resident; avg tuition + required fees'
    var_label(df_ipeds_pop[['tuitfee_md_nres']]) <- 'MD, full-time, non-resident; avg tuition + required fees'
    var_label(df_ipeds_pop[['tuitfee_law_res']]) <- 'Law, full-time, state resident; avg tuition + required fees'
    var_label(df_ipeds_pop[['tuitfee_law_nres']]) <- 'Law, full-time, non-resident; avg tuition + required fees'
    
  # COA variables
    var_label(df_ipeds_pop[['coa_grad_res']]) <- 'graduate, full-time, state resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'
    var_label(df_ipeds_pop[['coa_grad_nres']]) <- 'graduate, full-time, non-resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'
    var_label(df_ipeds_pop[['coa_md_res']]) <- 'MD, full-time, state resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'
    var_label(df_ipeds_pop[['coa_md_nres']]) <- 'MD, full-time, non-resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'
    var_label(df_ipeds_pop[['coa_law_res']]) <- 'Law, full-time, state resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'
    var_label(df_ipeds_pop[['coa_law_nres']]) <- 'Law, full-time, non-resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'

df_ipeds_pop %>% glimpse()


##########
########## Create data frame of generated variables, with each variable meant to represent the entire population
##########


num_obs <- 10000

# Generate normal distribution w/ custom mean and sd
set.seed(124)
norm_dist <- rnorm(n = num_obs, mean = 50, sd = 5)

# Generate right-skewed distribution
set.seed(124)
rskew_dist <- rbeta(n = num_obs, shape1 = 2, shape2 = 5)

# Generate left-skewed distribution
set.seed(124)
lskew_dist <- rbeta(n = num_obs, shape1 = 5, shape2 = 2)

# Generate standard normal distribution (default is mean = 0 and sd = 1)
set.seed(124)
stdnorm_dist <- rnorm(n = num_obs, mean = 0, sd = 1)  # equivalent to rnorm(10)

# Create dataframe
df_generated_pop <- data.frame(norm_dist, rskew_dist, lskew_dist, stdnorm_dist)

# drop individual objects associated with each variable
rm(norm_dist,rskew_dist,lskew_dist,stdnorm_dist)
rm(num_obs)


##########
########## Create sample versions of generated population data frame and IPEDS population data frame
##########

# create sample version of our generated data
  set.seed(124) # set seed so that everyone ends up with the same random sample
  
  df_generated_sample <- df_generated_pop %>% sample_n(size = 200)
  df_generated_sample %>% glimpse()


# create sample version of our ipeds data

  set.seed(124) # set seed so that everyone ends up with the same random sample
  
  df_ipeds_sample <- df_ipeds_pop %>% sample_n(size = 200) 
  
  # compare mean of coa_grad_res between population and sample
  mean(df_ipeds_pop$coa_grad_res, na.rm = TRUE)
  mean(df_ipeds_sample$coa_grad_res, na.rm = TRUE)
```

4. We want to know from the `df_ipeds_pop` data frame if students who attend schools in New England area (region 1 in Bureau of Economic Analysis, or the New England Region) have higher books and supplies cost (`books_supplies`) than students who attend schools in the Far West area (region 8 of Bureau of Economic Analysis, or the Far West Region). 

Please write out $H_0$ null hypothesis and $H_a$ alternative hypothesis:

#ANSWER KEY

$H_0: \mu_{{region_1}} = \mu_{{region_8}}$
- (in words): $H_0:$ the population mean books and supplies cost for students attending school in the New England area, $\mu_{{region_1}}$, is the same as the population mean books and supplies cost for students attending school in the Far West area ($\mu_{{region_8}}$).

Two-sided alternative hypothesis ($H_a$)

- $H_a: \mu_{{region_1}} \ne \mu_{{region_8}}$
- (in words): $H_a:$ the population mean books and supplies cost for students attending school in the New England area, $\mu_{{region_1}}$, is different than the population mean books and supplies cost for students attending school in the Far West area ($\mu_{{region_8}}$).

5. First conduct the two sample t-test by hand
```{r}
# ANSWER KEY
# 1. Calculate the Means
region_1 <- df_ipeds_pop %>% filter(region == 1)
region_8 <- df_ipeds_pop %>% filter(region == 8)

mean_r1 <- mean(region_1$books_supplies)
mean_r8 <- mean(region_8$books_supplies)

# 2. Calculate the Variance
var_r1 <- var(region_1$books_supplies)
var_r8 <- var(region_8$books_supplies)

# 3. Calculate the Standard Errors
n_r1 <- length(region_1$books_supplies)
n_r8 <- length(region_8$books_supplies)

SE <- sqrt(var_r1/n_r1 + var_r8/n_r8)
SE

# 4. Calculate the t-statistic
t_stat <- (mean_r1 - mean_r8) / SE
t_stat

# 5. Calculate degrees of freedom
# This can be done using the formula for unequal variances (Welch’s t-test)
df <- (var_r1/n_r1 + var_r8/n_r8)^2 / 
      ((var_r1^2 / (n_r1^2 * (n_r1 - 1))) + (var_r8^2 / (n_r8^2 * (n_r8 - 1))))
df

# 6. Calculate the p-value 
# We can find the p-value of the calculated t-statistic by using the pt() function in R
p_value <- 2 * pt(-abs(t_stat), df)
p_value

```

6. Now, perform with the `t.test()` function in R
```{r}
# ANSWER KEY
# Perform the two-sample t-test
t_test_result <- t.test(region_1$books_supplies, region_8$books_supplies, var.equal = FALSE)

# Display the result
t_test_result
```

7. What is the p-value? What can we say about it in relation to an alpha level of 0.05? 

**ANSWER KEY:**

The p-value was `0.537` which is much greater than the typical alpha level or significance level of 0.05. 

8. What can we conclude from our two sample t-test? 

Since our p-value of `0.537` is greater than `0.05`, we fail to reject the null hypothesis. Which means, there is no significant different between the population mean books and supplies cost for students attending school in the New England area, $\mu_{{region_1}}$ and the population mean books and supplies cost for students attending school in the Far West area ($\mu_{{region_8}}$).


# Render to pdf and submit problem set 

**Render to pdf** by clicking the "Render" button near the top of your RStudio window (icon with blue arrow) or drop down "File" and select "Render to PDF"

- Go to the [class website] (Need to fill in classwebsite) and under "Final", click on the "Final submission link"  
- Submit both .qmd and pdf files  
- Use this naming convention "lastname_firstname_ps#" for your .Rmd and pdf files (e.g. lee_belle_final.qmd & lee_belle_final.pdf)  
