---
title: "EDUC 152: Introduction to Quantitative Research in Education: Regression Analysis"
subtitle: "Spring 2021"
author: 
date: 
urlcolor: blue
output: 
  # pdf_document
  # word_document:
  #   toc: no
  #   toc_depth: '3'
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    fig_caption: true
    highlight: tango
    theme: default
    df_print: tibble

---

# Course information

```{r echo=FALSE, message=FALSE}
library(kableExtra)
library(tidyverse)

t <- c("Class website", "Class Zoom link", "Class Slack workspace")
l <- c("https://anyone-can-cook.github.io/educ152/", "https://ucla.zoom.us/j/91707509593", "https://app.slack.com/client/T01S30RQY2E/C01SVPEPHPT")

info <- tibble(t, l)

kable(info,col.names = c("Resource","Link"), escape = T) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"), full_width = F, position = "left")
```

# Course description

EDUC 152 introduces students to regression as a tool to answer questions about education. Regression is commonly used to answer questions about "association claims" -- the relationship between variables -- and "causal claims" -- the causal effect of one variable on another. However, using regression appropriately requires thoughtfulness about what kinds of questions regression can answer, about the assumptions regression relies on, about the limitations of our data, and about how particular variables (e.g., "race" and "gender") are incorporated into analyses. Otherwise, regression results may be biased and may reify rather than interrogate problematic ideas. Therefore, the course emphasizes learning fundamental concepts of regression analysis and how these concepts can be thoughtfully applied to address different kinds of questions about education. The course also trains students how to read and critically assess research that uses regression. ECUC 152 integrates theory and application using the R programming language. Students will be assessed through four substantive take-home assignments, including a final, capstone assignment.

# Instructor and teaching assistants

## Instructor

**Ozan Jaquette**  

- Pronouns: he/him/his
- Office: Moore Hall, Room 3038
- Email: [ozanj@ucla.edu](ozanj@ucla.edu)
- Office hours:
  - Zoom office hours: Wednesdays 3 - 4 pm, zoom [link](https://ucla.zoom.us/j/2545750562)
  - And by appointment (afternoon preferred)
  
## Teaching assistants

**Patricia Martín**  

- Pronouns: she/her/hers/ella
- Email: [pmarti@g.ucla.edu](pmarti@g.ucla.edu)
- Office hours:
  - Zoom office hours: Thursdays 12 - 1pm, zoom [link](
 https://ucla.zoom.us/j/97786470715)
  - And by appointment

# Course learning goals

__Big-picture (conceptual) learning goals__

1. Apply principles of statistical inference to state and test hypotheses
1. Know the fundamental concepts of multiple regression analysis and the assumptions that must be satisfied to make inferences from regression results
1. Interpret regression results and communicate these results in non-technical language
1. Understand the principles of causal inference, including why experiments yield unbiased estimates of causal effects and how to use regression to answer causal research questions
1. Understand ethical concerns and problematic practices common in regression-based research, be able to identify these concerns and practices in empirical research, and reflect on how you want to approach these issues in your own work
1. Understand and critically evaluate empirical research that uses regression to answer research questions

__Skill-based learning goals__

1. Estimate linear regression models using R
1. Develop proficiency in basic data management skills, including simple descriptive statistics to investigate data quality and creating analysis variables
1. Create documents using RMarkdown and Rstudio that contain research results, including: graphical visualizations; tables of descriptive statistics and regression results; and in-text citations and APA formatted reference lists

# Extended course description and how to succeed

Regression is a quantitative methodology commonly used to answer questions about the “association” between variables and questions about the causal effect of one variable on another.  Association research questions examine the relationship between variables. For example, what is the relationship between property values and school expenditure per student. However, the existence of an association relationship does not necessarily mean that the value of one variable affects the value of the other. For example, we would expect to observe a positive relationship between ice cream sales and sunglasses sales in a city, but it is unlikely that ice cream sales have a positive causal effect on sunglasses sales. Rather, the temperature and intensity of the sun are likely driving sales of both ice cream and sunglasses. Causal research questions attempt to isolate the causal effect of one variable on another and can usually be stated in the form, “what is the effect of X on Y.” For example, what is the effect of reducing the number of students per class on student learning outcomes?

This course teaches the fundamental principles of regression analysis and the application of these principles to questions about education. Using regression appropriately requires thoughtfulness about what kinds of questions regression can answer, about the assumptions regression relies on, about the limitations of our data, and about how particular variables (e.g., "race" and "gender") are incorporated into analyses. Therefore, the course emphasizes how the principles of regression can be thoughtfully applied to analyze association research questions and causal research questions. 

The course integrates statistical theory and application using the _R_ programming language. The course consists of three units: first, the basics of statistical inference necessary for hypothesis testing; second, the fundamental principles of regression, applied to questions of association; and, third, using regression to answer causal research questions.

The primary course assessments are four "problem sets." Each problem set will require students to apply knowledge of statistical concepts, and conduct substantive statistical analyses around a particular research question. Students will complete the first three problem sets in groups. Students will complete the final capstone problem set, due during finals week, on their own.

## Course structure

Each week, the course will be structured around asynchronous (pre-class) lectures and one synchronous workshop-style class meeting per week. Weekly homework will consist of students working through the lectures on their own and a modest amount of required reading.

<!-- -
Students will work through asynchronous video lectures and lectures slides on their own. These lectures introduce statistical theory, the relevant programming skills, and provide  code and real-world data so that students can practice running code on their own computer. 

During synchronous class time students will work in groups to practice a new skill (e.g., creating .Rmd files with in-text citations and references), solve a practical research challenge (e.g., create analysis variables from survey data), or deconstruct empirical research articles we read.
-->



1. __Asynchronous (pre-class) lectures__. Weekly asynchronous lectures will be posted on the course website with the expectation that students work through the lecture in advance of our weekly synchronous class meeting. Lecture materials will consist of three types of resources: first, detailed lecture slides (PDF or HTML) introducing the statistical theory, programming skills, and sample code; second, short videos (e.g., 15 to 30 total minutes per week) that provide a high-level discussion of important and/or challenging concepts from the lecture slides, but not a line-by-line recitation of the lecture; and, third, the .Rmd file that created the PDF/HTML lecture slides. This .Rmd file will contain all "code chunks" and links to all data utilized in the lecture. Thus, students will be able to "learn by doing" in that they can run R code on their own computer while they work through lecture materials on their own.
1. __Synchronous workshop-style class meetings.__ We will have one synchronous class meeting per week. Typically, these meetings will begin with a discussion of concepts students found difficult or confusing from the lecture materials. The majority of class time will be devoted to students working in groups on a substantive activity posed by the instructor, for example: practicing a new skill, like creating .Rmd files with in-text citations and references; solving a practical research challenge, like creating analysis variables from survey data in the presence of skip patterns; or deconstructing and critiquing empirical research articles we read prior to class. While students work in groups, the instructor and TA will visit each group to answer questions and talk through ideas.

## How to succeed in this class

Prior to our in-class meetings, students should work through lecture materials on their own. We recommend treating the lecture materials as an active learning experience, in which students run R code on their computer instead of merely reading text on the slide. Additionally, we recommend that students ask questions on Slack when they are having difficulty with the material.

With respect to written work, the problem sets -- described below -- will be substantive and are intended to be challenging. Students who devote time each week working through the lecture materials will be better prepared for the problem sets. We recommend starting the problem sets early. This way students will have plenty of time to ask for help on questions they find challenging.

Concepts introduced earlier in the quarter will be utilized later in the quarter. For example, in the beginning of the quarter we learn how to test hypotheses about descriptive claims (e.g., test a hypothesis about average household income). Later in the quarter we will utilize these same concepts to test hypotheses about regression models. Therefore, please seek help if you feel confused about new concepts introduced in a given week so that you do not feel confused when these concepts reappear later in the quarter.

## Prerequisites

The prerequisites for EDUC 152 are:

- EDUC 150 (Introduction to Quantitative Research in Education: Claims and Evidence) __OR__ EDUC 151 (Introduction to Quantitative Research in Education: Introduction to Measurement and Assessment in Education)
- Basic familiarity with _R_, the programming language and software environment we will use

Students who have not completed EDUC 150 or EDUC 151 but who have completed a different introductory statistics course can enroll in EDUC 152 with instructor consent.

# Classroom environment

We all have a responsibility to ensure that every member of the class feels valued and safe. 
Be mindful that our words and body language affects others in ways we might not fully understand. We have a responsibility to express our ideas in a way that doesn't make disparaging generalizations and doesn't make people feel excluded. As an instructor, I am responsible for setting an example through my own conduct.

Learning regression, while trying to get a handle on R and unfamiliar data can feel overwhelming! We must create an environment where students feel comfortable asking questions and talking about what they did not understand. Discomfort is part of the learning process. Unburdern yourself from the weight of being an "expert." Focus your energy on improving and helping your classmates improve.

##  Towards an anti-racist learning experience

Every course should be an anti-racist course, even when the subject matter is broadly oriented. In this course we'll engage with research that reflect systemic gaps based on race, ethnicity, immigration status, and gender identity, among other aspects of identity. We will discuss whether the language, the framing, the analyses, and even the research question of a study contribute to problematic beliefs. If so, how can we do better? It is also critical that we acknowledge that the social and economic marginalization reflected in data is rooted in systemic oppression that upholds opportunity for some at the expense of others. We should all be thinking about our own role in upholding these systems.


# Course website and communication

## Course website

All course related material can be found on the [course website](https://anyone-can-cook.github.io/educ152). Pre-recorded lecture videos, lecture slides (PDF, HTML), and .Rmd files will be posted on the class website under the associated sections. Additional resources (e.g., syllabus) may also be posted on the class website. 

## Course communication and online discussion

Instead of communicating via email, we’ll use *Slack*, an online collaborative workspace. You can access it through a web browser or by downloading the Slack app on your phone or computer. Our workspace is called "EDUC 152, Spring 2021" and the link is at the top of the syllabus. The best way to communicate with the instructor, TAs, and your classmates is via Slack, instead of email. All class announcements (e.g., "the due date for the assignment has changed" or "I've posted a study guide – check it out!") will be made with Slack. This means it’s important that you check regularly for communication.

If you have a personal question or issue, please send a direct message to the instructor and/or TA via Slack. Additionally, we are available for office hours or by appointment if there is anything you would like to discuss with us in private. 

# Course materials

**Required book to buy (or get from library)**

- Stock, J., and Watson, M. _Introduction to econometrics_ (1st, 2nd, 3rd, 4th editions are fine; just not the "brief edition")
  - This will be our primary text for learning regression
  - [LINK](https://www.amazon.com/Introduction-Econometrics-Stock-James-Watson-dp-935286350X/dp/935286350X/ref=mt_other?_encoding=UTF8&me=&qid=) to paperback 3rd edition on Amazon

  
**Free, online books/resources we will use for assigned reading**  

- Lane, D.M. (2020) _Online statistics education: A multimedia course of study_
  - [Link to onlinestatbook.com](https://onlinestatbook.com/index.html)
  - This will be our primary text for learning about sampling distribution and hypothesis testing at the beginning of the course
- Hanck, Arnold, Gerber, and Schmelzer (2020). _Introduction to econometrics with R_.
  - [Link to Introduction to econometrics with R](https://www.econometrics-with-r.org/)
  - This book is about regression analysis using _R_ and is written as "an empirical companion" to _Introduction to econometrics_ by Stock and Watson


**Other free resources we may draw from**

- [R Markdown: The Definative Guide](https://bookdown.org/yihui/rmarkdown/) by Yihui Xie, J. J. Allaire, and Garrett Grolemund
- [R for Data Science](https://r4ds.had.co.nz/) by Garrett Grolemund and Hadley Wickham
- Two-course programming/data science sequence (using _R_) developed by Ozan Jaquette, Patricia Martín, and Crystal Han
  - [Link to Rclass1](https://anyone-can-cook.github.io/rclass1/), EDUC260A: Introduction to programming and data management
  - [Link to Rclass2](https://anyone-can-cook.github.io/rclass2/), EDUC260B: Fundamentals of programming


Links to other required and optional reading will be on the [course website](https://anyone-can-cook.github.io/educ152/).


**Required software**

- **_[R](https://cloud.r-project.org/)_**, statistical programming language [**FREE!**]
- **[RStudio](https://rstudio.com/products/rstudio/download/)**, integrated development environment for R  [**FREE!**]


Link to tips for software installation [HERE](https://github.com/anyone-can-cook/educ152/raw/main/assets/resources/software_installation.pdf).

- **_Note_**: _This document is subject to change prior to spring quarter._

# Assignments and grading

Course grade will be based on the following components:

- Problem set 1, 15%
- Problem set 2, 15%
- Problem set 3, 15%
- Short exercise, 5%
- Problem set 4, 25%
- In-class group activities, 15%
- Attendance and participation, 10%

## Problem sets & short exercise (75 percent of total grade)

The primary course assessments are four problem sets and a short exercise. The first three problem sets are each worth 15% of the course grade and students will work in groups. Additionally, there is a short exercise worth 5% of the course grade. The final, capstone problem set is worth 30% and students will work alone. 

Each problem set will require students to apply knowledge of statistical concepts, conduct substantive statistical analyses, and present and interpret results. Other questions will introduce students to some of the thorny data challenges that inevitably arise in real research projects. The capstone problem set will require students to conduct the major components of an empirical regression analysis, from research question and variable collection to modeling, presentation, and interpretation. Additionally, the capstone problem set will require students to critically evaluate an empirical journal article that utilized the same data sources to answer the same research question.


You will work in groups for the first three problem sets. **However, it is important that you understand how to do the problem set on your own, rather than copying the solution developed by group members.**  Each student will submit their own R script or .Rmd file. Since you will be working together, it is understandable that answers for some questions will be the same as your group members. However, if I find compelling evidence that a student merely copied solutions from a classmate, I will consider this a violation of academic integrity and that student will receive a zero for the homework assignment.

Late submissions will lose 20% (i.e., max grade becomes 80%). Problem sets submitted a week late (after the problem set is due) will not receive points. You will not lose points for late submission if you cannot submit a problem set due to an unexpected emergency. Extensions can be granted, but only with prior approval by the instructor(s). But please contact the instructor by email as soon as you can so we can work out a plan.

We strongly recommend using the course Slack workspace to ask questions that you have about problem sets. Instructors will do our best to reply quickly with helpful hints/explanations and we encourage members of the class to do the same.

Link to problem set expectations and helpful resources [HERE](https://github.com/ozanj/rclass/raw/master/lectures/problemset_resources.pdf).

- **_Note_**: _This document is subject to change prior to spring quarter._

<!-- -

- FOR CDAS SAY PROBLEM SETS ARE NOT FULLY DEVELOPED BUT HERE IS AN EXAMPLE OF A SIMILAR PROBLEM SET
- 

COMMENTS FROM THE UNDERGRAD STATS COMMITTEE

- recommmendation on how challenging
  - make first problem set credit/no credit
  - make them short but challenging;
  - take time in class to work on the problem set
  - make the first couple of problem sets that have a lot of hints;
- think about how grades would be distributed throughout; you can't give everyone an A;
- maybe giving students an opportunity to revise?
- Rashmita: Lucrecia's example of "ED471-Project 3 - Spring 2019" problem set is  problem set is 

-->

## In-class group activities (15 percent of total grade)

During most synchronous class sessions, students will work on a group activity or challenge that applies concepts and skills that we are learning. For example, you may be asked to run a statistical test in R and use the test results object to create a formatted table that you insert into a .Rmd file. Or, you might write a draft critique of the methodology of a published research paper. Students will submit their work at the end of class. Most of these tasks cannot be fully completed during the duration of the class. The goal is to get students thinking. Students will be graded largely based on effort.

## Attendance and participation (10 percent of total grade)

Students are expected to participate in the weekly class meetings by being attentive, supportive, by asking questions, or by answering questions posed by others on Zoom. Additionally, students can receive strong participation grades by asking questions and answering questions posed by classmates on Slack. 


## Grading scale

```{r echo=FALSE, message=FALSE}
library(kableExtra)
library(tidyverse)

lg <- c("A+", "A", "A-", "B+", "B", "B-", "C+", "C", "C-", "D", "F")
nv <- c("99-100%", "93<99%", "90<93%", "87<90%", "83<87%", "80<83%", "77<80%", "73<77%", "70<73%", "60<70%", "0<60%")

grade <- tibble(lg, nv)

kable(grade,col.names = c("Letter Grade", "Percentage"), escape = F) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left")
  
```


# Course schedule

```{r, eval = FALSE, include = FALSE}
1. Understand principles of statistical inference for hypothesis testing, including:
    - Population parameters vs. point estimates
    - Properties of sampling distributions
    - Standard error and central limit theorem
    - Developing formal hypotheses and hypothesis testing
1. Develop a deep, applied understanding of multivariate regression, including:
    - The population regression model
    - Ordinary least squares estimator of the population regression model
    - Prediction and measures of model fit
    - Hypothesis testing and confidence intervals about the relationship between an independent variable, X, and the outcome, Y
    - Assumptions that must be satisfied in order to make inferences based on regression
1. Understand principles of causal inference, including
    - Counterfactuals and Rubin’s causal model
    - Why are experiments the gold standard of causal inference
    - Why regression analysis of observational (non-experimental data) is unlikely to yield unbiased estimates of causal effects (e.g., endogeneity, exogenous variation, omitted variable bias)
    - Using regression to estimate correlational relationships vs. causal relationships
1. Critically assess empirical research that uses regression, including:
    - How to read and understand methods sections and results sections of empirical research
    - Critique the methodological decisions of empirical research, assess whether the research provides a sound evidentiary basis for claims they make

__Skill-based learning goals__

1. Conduct regression analysis using R, including:
    - Estimate linear regression models and interpret regression results
      - Models using a continuous or categorical outcome variable
      - Models that include continuous or categorical independent variables
1. Develop proficiency in basic data management skills, including
    - Using simple descriptive statistics to investigate variables prior to analysis
    - Creating analysis variables
1. Use R/Rstudio/RMarkdown to produce documents containing research results, including
    - Creating papers and policy reports
    - inserting in-text references and and reference lists
    - Create clear tables that summarize regression results
    - Creating tables and graphs of descriptive statistics

```
Below is an overview of the tentative course schedule, which is subject to change at the discretion of the instructor. Topics may be cut if we need to devote more time to learning the most central topics. It is unlikely that additional topics will be added. The official course schedule, required reading, and optional reading will be posted on the [course website](https://anyone-can-cook.github.io/educ152).

**Problem set distribution and due dates are tentative and may be subject to change** 


__UNIT 1: FUNDAMENTALS OF STATISTICAL INFERENCE__

<span style="background-color: #CC76AD"> <b>WEEK 1 (04/02/21) </b></span>

- Lecture topics  
    - Course overview; Goal of statistics; Review of basic statistics concepts (e.g. standard deviation)
    - *Asynchronous lecture posted **(Friday night on 3/26)***  


<span style="background-color: #CC76AD"> <b>WEEK 2 (04/09/21) </b></span>

- Lecture topics  
    - Distributions; Sampling distribution; Central limit theorem  
    - *Asynchronous lecture posted **(Friday night on 4/2)***  
- **Problem set 1 distributed 04/09**


<span style="background-color: #CC76AD"> <b>WEEK 3 (04/16/21) </b></span>

- Lecture topics  
    - Hypothesis testing; Hypothesis testing about a single population mean
    - *Asynchronous lecture posted **(Friday night on 4/9)***   
- **Problem set 1 due 04/16**  

 
<span style="background-color: #CC76AD"> <b>WEEK 4 (04/23/21) </b></span>

- Lecture topics  
    - Fundamental concepts in causal inference; Why experiments work; Hypotheses comparing population means of two groups  
    - *Asynchronous lecture posted **(Friday night on 4/16)***
- **Problem set 2 distributed 04/23**

__UNIT 2: FUNDAMENTALS OF REGRESSION__  

<span style="background-color: #CC76AD"> <b>WEEK 5 (04/30/21) </b></span>

- Lecture topics  
    - Introduction to bivariate regression 
    - *Asynchronous lecture posted **(Friday night on 4/23)***  
- **Problem set 2 due 04/30**


<span style="background-color: #CC76AD"> <b>WEEK 6 (05/07/21) </b></span>

- Lecture topics  
    - Prediction/OLS prediction line; Measures of model fit
    - *Asynchronous lecture posted **(Friday night on 4/30)*** 
- **Problem set 3 distributed 05/07**


<span style="background-color: #CC76AD"> <b>WEEK 7 (05/14/21) </b></span>

- Lecture topics  
    - Hypothesis testing and confidence intervals about Bhat; (also, interpretation of Bhat)
    - *Asynchronous lecture posted **(Friday night on 5/7)***
- **Problem set 3 due 05/14**


<span style="background-color: #CC76AD"> <b>WEEK 8 (05/21/21) </b></span>

- Lecture topics  
    - Categorical X variables; Introduction to multivariate regression
    - *Asynchronous lecture posted **(Friday night on 5/14)***  
- **Short exercise distributed 05/14**

<span style="background-color: #CC76AD"> <b>WEEK 9 (05/28/21) </b></span>
  
- Lecture topics  
    - Reading and assessing research that uses regression  
    - *Asynchronous lecture posted **(Friday night on 5/21)***
- **Short exercise due 05/28**
- **Final Problem set distributed 05/28**


<span style="background-color: #CC76AD"> <b>WEEK 10 (06/04/21) </b></span>

- Lecture topics  
    - Using regression for causal inference; OLS assumptions and omitted variable bias
    - *Asynchronous lecture posted **(Friday night on 5/28)***  
- **Final Problem set due 06/11**



# Course policies

## Learning during a global pandemic

With the ongoing spread of the COVID-19 pandemic, we understand that right now is a challenging time for everybody. Many of us may be experiencing added stress or responsibilities that make learning and completing classwork difficult. If you are having trouble keeping up with the class, please reach out to the teaching team and we will help work out a plan with you. We understand that right now is a precarious time and in the event that you or someone in your family and/or shared living space gets sick, we ask that you please reach out to us as soon as you are able to. We want to be accommodating to everyone's unique situation and hope to make this class an enjoyable learning experience for all.

## Online collaboration/netiquette

You will communicate with instructors and peers virtually through a variety of tools such as Slack, email, and Zoom web conferencing. The following guidelines will enable everyone in the course to participate and collaborate in a productive, safe environment.

-	Be professional, courteous, and respectful as you would in a physical classroom.
-	Online communication lacks the nonverbal cues that provide much of the meaning and nuances in face-to-face conversations. Choose your words carefully, phrase your sentences clearly, and stay on topic.
-	It is expected that students may disagree with the research presented or the opinions of their fellow classmates. To disagree is fine but to disparage others' views is unacceptable. All comments should be kept civil and thoughtful.
- It is imperative that we respect one another in this course, and all other spaces. One way to gain/show respect is to actively listen to one another. Please do not text, tweet, email, Facebook, LinkedIn, browse the internet, and such during class.
- In the unlikely event that Zoom is down, please be sure to check the class Slack channel often for instructions on how we will complete that class session in an asynchronous manner.

**Class Zoom guidelines**  

All synchronous class sessions will be held online, via Zoom. Below, we have outlined some general guidelines about Zoom learning. As we continue learning together, we can add to and change the below list. I'm open to your feedback and your experiences as we continue to learn how to learn via Zoom.  

- **Video**: Students are **strongly** encouraged to turn on their video during synchronous lectures. Additionally, we encourage students to turn on their video particularly during small group breakout rooms. You may turn off your camera in the event that you have to step out for a few minutes and/or are having connectivity issues.   
- **Audio**: We ask students to mute their microphones when they are not speaking. We encourage the use of earphones or headphones if you are in a space with background noise.  
- **Zoom outage**: In the unlikely event that Zoom is down, the instructors will post an announcement on Slack with instructions for completing the class section in an asynchronous manner. Therefore, if Zoom is not functioning properly during the class period, be sure to check the class Slack channel.  
- **Internet connectivity**: We understand that having access to a stable internet connection and/or electronic equipment is a privilege. With that in mind, we want to provide a space where everyone has the resources they need to do well in the class. If you have any issues with your internet connection and/or don't have access to electronic equipment, please reach out to the instructors. 

## Academic accomodations

**Center for Accessible Education**  

Students needing academic accommodations based on a disability should contact the Center for Accessible Education (CAE). When possible, students should contact the CAE within the first two weeks of the term as reasonable notice is needed to coordinate accommodations. For more information visit https://www.cae.ucla.edu/.

Located in A255 Murphy Hall: (310) 825-1501, TDD (310) 206-6083; http://www.cae.ucla.edu/

-	Due to COVID-19, the CAE office is closed for in-person meetings
-	CAE counselor, resources, and services are still available via email / virtual appointment/drop-in hours
-	Stay up-to-date with CAE newsletters & announcements at https://www.cae.ucla.edu/announcements-events/student



## Academic integrity

__UCLA policy__  

  - UCLA is a community of scholars. In this community, all members including faculty, staff and students alike are responsible for maintaining standards of academic honesty. As a student and member of the University community, you are here to get an education and are, therefore, expected to demonstrate integrity in your academic endeavors. You are evaluated on your own merits. Cheating, plagiarism, collaborative work, multiple submissions without the permission of the professor, or other kinds of academic dishonesty are considered unacceptable behavior and will result in formal disciplinary proceedings.  
  
__This class__  

  - Given that 75% of course grade is based on problem sets, the primary academic honesty concern that could come up in this class is copying problem set solutions from somebody else and passing this in as your own work.


# Campus resources


## Counseling and Psychological Services (CAPS)

As a student you may experience a range of issues that can cause barriers to learning, such as strained relationships, increased anxiety, alcohol/drug problems, depression, difficulty concentrating and/or lack of motivation. These mental health concerns or stressful events may lead to diminished academic performance or reduce a student's ability to participate in daily activities. UC offers services to assist you with addressing these and other concerns you may be experiencing. If you or someone you know are suffering from any of the aforementioned conditions, consider utilizing the confidential mental health services available on campus.

Students in distress may speak directly with a counselor 24/7 at (310) 825-0768, or may call 911; located in Wooden Center West; https://www.caps.ucla.edu

-	CAPS is open and has transitioned to Telehealth services ONLY
-	Open Mon – Thurs: 8am-6pm and Fri: 8am-5pm
-	As always, 24/7 crisis support is always available by phone at (310) 825-0768


## Discrimination

UCLA is committed to maintaining a campus community that provides the stronget possible support for the intellectual and personal growth of all its members- students, faculty, and staff. Acts intended to create a hostile climate are unacceptable.  

  - To file an online incident report, visit: https://equity.ucla.edu/report-an-incident/  


## LGBTQ resource center

The LGBTQ resource center provides a range of education and advocacy services supporting intersectional identity development. It fosters unity; wellness; and an open, safe, inclusive environment for lesbian, gay, bisexual, intersex, transgender, queer, asexual, questioning, and same-gender-loving students, their families, and the entire campus community. Find it in the Student Activities Center, or via email lgbt\@lgbt.ucla.edu.  

  - Visit their website for more information: https://www.lgbt.ucla.edu/ and virtual upcoming events

## International students

The Dashew Center provides a range of programs to promote cross-cultural learning, language improvement, and cultural adjustment. Their programs include trips in the LA area, performances, and on-campus events and workshops.  

  -	Due to COVID-19, the Dashew Center has transitioned its operations to a remote setting
  - Visit their website for more information: https://www.internationalcenter.ucla.edu/  
  - For COVID updates, visit https://www.internationalcenter.ucla.edu/covid-19-updates

## UCLA Undocumented Student Program

This program provides a safe space for undergraduate and graduate undocument students. USP supports the UndocuBruin community through personalized services and resources, programs, and workshops.  

  - Visit their website for more information: https://www.usp.ucla.edu/  
  - You can reach USP at usp@saonet.ucla.edu

## Student legal services

UCLA Student Legal Services provides a range of legal support to all registered and enrolled UCLA students. Some of their services include:  

  - Landlord/Tenant Relations (Including challenges during COVID)
  - Accident and Injury Problems
  - Domestic Violence and Harassment
  - Divorces and Other Family Law Matter  

Due to COVID, Student legal Services is closed to walk-ins.  

  - All services are by appointment only
  - For more information visit their website: http://www.studentlegal.ucla.edu/index.php  
  

## Students with Dependents

UCLA Students with Dependents provides support to UCLA studens who are parents, guardians, and caregivers. Some of their services include:  

- Information, referrals, and support to navigate UCLA (childcare, family housing, financial aid)
- Access to information about resources within the larger community
- On-site application and verification for CalFresh (food stamps) & MediCal and assistance with Cal Works/GAIN
- A quiet study space
- Family friendly graduation celebration in June  

For more information visit their website: https://www.swd.ucla.edu/

## Campus maps 

**Lactation Rooms** 

- [Map to lactation rooms on campus](https://ucla.app.box.com/v/2019-lactation-map)  

**Gender Inclusive restrooms**  

- [Map to gender inclusive restrooms](https://www.lgbt.ucla.edu/Portals/38/Documents/GIRR%20Map%2007-2019%20Compressed.pdf)  

**Campus accessibility**  

- [Campus accessibility map](http://map.ucla.edu/downloads/pdf/Access_08_21_15.pdf)


## Title IX Resources

Title IX prohibits gender discrimination, including sexual harassment, domestic and dating violence, sexual assault, and stalking.  If you have experienced sexual harassment or sexual violence, there are a variety of resources to assist you.  

- __CONFIDENTIAL RESOURCES__:You can receive confidential support and advocacy at the CARE Advocacy Office for Sexual and Gender-Based Violence, A233 Murphy Hall, CAREadvocate@careprogram.ucla.edu, (310) 206-2465. Counseling and Psychological Services (CAPS) also provides confidential counseling to all students and can be reached 24/7 at (310) 825-0768.  

- __NON-CONFIDENTIAL RESOURCES__: You can also report sexual violence or sexual harassment directly to the University's Title IX Coordinator, 2255 Murphy Hall, titleix\@conet.ucla.edu, (310) 206-3417. Reports to law enforcement can be made to UCPD at (310) 825-1491. These offices may be required to pursue an official investigation.

*Faculty and TAs are required under the UC Policy on Sexual Violence and Sexual Harassment to inform the Title IX Coordinator should they become aware that you or any other student has experienced sexual violence or sexual harassment.*  

## Undergraduate Writing Center

Peer learning facilitators (PLFs) are undergraduates who understand the challenges of writing at UCLA. Scheduled appointment and walk-in options are available, see https://uwc.ucla.edu/ for more information about writing programs and to get assistance with your writing.

-	Due to COVID-19, all physical UWC offices will be closed until further notice
-	All UWC appointments are now online via Zoom and Google Docs
-	Spring 2021 virtual drop-in appointments M 10am-12pm & 2-4pm | Tues. 1-3pm | W 3-5pm | Th 10am-12pm & 3-5pm | F 10am-12pm &1-3pm

For additional campus resources and student services, please review [this document](https://ucla.app.box.com/s/st51pify842ouoc24ae2r80s9sesavmy). 
