---
title: "Social Sciences Intro to Statistics"
subtitle: "Week 10.1 Introduction to Multivariate Regression"
format: pdf
editor: source
---
Week 10: Learning goal - Test multivariate regression and interpret findings to estimate causal effects. 

 
```{r, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", highlight = TRUE, warning = FALSE, message = FALSE)
  #comment = "#>" makes it so results from a code chunk start with "#>"; default is "##"
```

# Introduction

Lecture Overview: 

- Multivariate regression model
- Interpretation

Load packages:
```{r}
library(tidyverse)
library(ggplot2)
library(labelled)
library(patchwork)

# Load ipeds dataset from course website
load(url('https://raw.githubusercontent.com/bcl96/Social-Sciences-Stats/main/data/ipeds/output_data/panel_data.RData'))
```


```{r, echo=FALSE}
# Create ipeds data frame with fewer variables/observations
df_ipeds_pop <- panel_data %>%
  # keep data from fall 2022
  filter(year == 2022) %>%
  # which universities to keep:
    # 2015 carnegie classification: keep research universities (15,16,17) and master's universities (18,19,20)
  filter(c15basic %in% c(15,16,17,18,19,20)) %>%
  # which variables to keep
  select(instnm,unitid,opeid6,opeid,control,c15basic,stabbr,city,zip,locale,obereg, # basic institutional characteristics
         tuition6,fee6,tuition7,fee7, # avg tuition and fees for full-time grad, in-state and out-of-state
         isprof3,ispfee3,osprof3,ospfee3, # avg tuition and fees for MD, in-state and out-of-state
         isprof9,ispfee9,osprof9,ospfee9, # avg tuition and fees for Law, in-state and out-of-state
         chg4ay3,chg7ay3,chg8ay3) %>% # [undergraduate] books+supplies; off-campus (not with family) room and board; off-campus (not with family) other expenses
  # rename variables; syntax <new_name> = <old_name>
  rename(region = obereg, # revion
         tuit_grad_res = tuition6, fee_grad_res = fee6, tuit_grad_nres = tuition7, fee_grad_nres = fee7, # grad
         tuit_md_res = isprof3, fee_md_res = ispfee3, tuit_md_nres = osprof3, fee_md_nres = ospfee3, # md
         tuit_law_res = isprof9, fee_law_res = ispfee9, tuit_law_nres = osprof9, fee_law_nres = ospfee9, # law
         books_supplies = chg4ay3, roomboard_off = chg7ay3, oth_expense_off = chg8ay3) %>% # [undergraduate] expenses
  # create measures of tuition+fees
  mutate(
    tuitfee_grad_res = tuit_grad_res + fee_grad_res, # graduate, state resident
    tuitfee_grad_nres = tuit_grad_nres + fee_grad_nres, # graduate, non-resident
    tuitfee_md_res = tuit_md_res + fee_md_res, # MD, state resident
    tuitfee_md_nres = tuit_md_nres + fee_md_nres, # MD, non-resident
    tuitfee_law_res = tuit_law_res + fee_law_res, # Law, state resident
    tuitfee_law_nres = tuit_law_nres + fee_law_nres) %>% # Law, non-resident  
  # create measures of cost-of-attendance (COA) as the sum of tuition, fees, book, living expenses
  mutate(
    coa_grad_res = tuit_grad_res + fee_grad_res + books_supplies + roomboard_off + oth_expense_off, # graduate, state resident
    coa_grad_nres = tuit_grad_nres + fee_grad_nres + books_supplies + roomboard_off + oth_expense_off, # graduate, non-resident
    coa_md_res = tuit_md_res + fee_md_res + books_supplies + roomboard_off + oth_expense_off, # MD, state resident
    coa_md_nres = tuit_md_nres + fee_md_nres + books_supplies + roomboard_off + oth_expense_off, # MD, non-resident
    coa_law_res = tuit_law_res + fee_law_res + books_supplies + roomboard_off + oth_expense_off, # Law, state resident
    coa_law_nres = tuit_law_nres + fee_law_nres + books_supplies + roomboard_off + oth_expense_off) %>% # Law, non-resident    
  # keep only observations that have non-missing values for the variable coa_grad_res
    # this does cause us to lose some interesting universities, but doing this will eliminate some needless complications with respect to learning core concepts about statistical inference
  filter(!is.na(coa_grad_res))

# Add variable labels to the tuit+fees variables and coa variables
  # tuition + fees variables
    var_label(df_ipeds_pop[['tuitfee_grad_res']]) <- 'graduate, full-time, resident; avg tuition + required fees'
    var_label(df_ipeds_pop[['tuitfee_grad_nres']]) <- 'graduate, full-time, non-resident; avg tuition + required fees'
    var_label(df_ipeds_pop[['tuitfee_md_res']]) <- 'MD, full-time, state resident; avg tuition + required fees'
    var_label(df_ipeds_pop[['tuitfee_md_nres']]) <- 'MD, full-time, non-resident; avg tuition + required fees'
    var_label(df_ipeds_pop[['tuitfee_law_res']]) <- 'Law, full-time, state resident; avg tuition + required fees'
    var_label(df_ipeds_pop[['tuitfee_law_nres']]) <- 'Law, full-time, non-resident; avg tuition + required fees'
    
  # COA variables
    var_label(df_ipeds_pop[['coa_grad_res']]) <- 'graduate, full-time, state resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'
    var_label(df_ipeds_pop[['coa_grad_nres']]) <- 'graduate, full-time, non-resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'
    var_label(df_ipeds_pop[['coa_md_res']]) <- 'MD, full-time, state resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'
    var_label(df_ipeds_pop[['coa_md_nres']]) <- 'MD, full-time, non-resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'
    var_label(df_ipeds_pop[['coa_law_res']]) <- 'Law, full-time, state resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'
    var_label(df_ipeds_pop[['coa_law_nres']]) <- 'Law, full-time, non-resident COA; == tuition + fees + (ug) books/supplies + (ug) off-campus room and board + (ug) off-campus other expenses'

df_ipeds_pop %>% glimpse()


##########
########## Create data frame of generated variables, with each variable meant to represent the entire population
##########


num_obs <- 10000

# Generate normal distribution w/ custom mean and sd
set.seed(124)
norm_dist <- rnorm(n = num_obs, mean = 50, sd = 5)

# Generate right-skewed distribution
set.seed(124)
rskew_dist <- rbeta(n = num_obs, shape1 = 2, shape2 = 5)

# Generate left-skewed distribution
set.seed(124)
lskew_dist <- rbeta(n = num_obs, shape1 = 5, shape2 = 2)

# Generate standard normal distribution (default is mean = 0 and sd = 1)
set.seed(124)
stdnorm_dist <- rnorm(n = num_obs, mean = 0, sd = 1)  # equivalent to rnorm(10)

# Create dataframe
df_generated_pop <- data.frame(norm_dist, rskew_dist, lskew_dist, stdnorm_dist)

# drop individual objects associated with each variable
rm(norm_dist,rskew_dist,lskew_dist,stdnorm_dist)
rm(num_obs)


##########
########## Create sample versions of generated population data frame and IPEDS population data frame
##########

# create sample version of our generated data
  set.seed(124) # set seed so that everyone ends up with the same random sample
  
  df_generated_sample <- df_generated_pop %>% sample_n(size = 200)
  df_generated_sample %>% glimpse()


# create sample version of our ipeds data

  set.seed(124) # set seed so that everyone ends up with the same random sample
  
  df_ipeds_sample <- df_ipeds_pop %>% sample_n(size = 200) 
  
  # compare mean of coa_grad_res between population and sample
  mean(df_ipeds_pop$coa_grad_res, na.rm = TRUE)
  mean(df_ipeds_sample$coa_grad_res, na.rm = TRUE)

##########
########## STAR DATA
##########

# load star data
load(url("https://raw.githubusercontent.com/bcl96/Social-Sciences-Stats/main/data/star/star_panel_data.RData"))

#df_star_panel %>% glimpse()

# create data frame for STAR experiment, keeping only kindergarten
df_stark <- df_star_panel %>% 
  # keep only kindergarten year
  filter(grade ==1) %>% 
  # keep only observations with non-missing value for reading score
  filter(!is.na(read)) %>%
  # keep only observations with non-missing values for treatment assignment
  filter(!is.na(star)) %>%
  # drop observations where treatment status is regular+aide
  filter(star !=3) %>%
  # keep selected variables
  select(id,grade,star,read,gender,ethnicity,lunch,school,degree,experience) %>%
  # create a variable "treatment" that equals 1 if student receives treatment (small class) and equals 0 otherwise
  mutate(
    treatment = if_else(star==2,1,0)
  )


df_stark %>% glimpse()

```

### Multivariate Regression Model

For the multivariate regression lecture, we will focus on using multivariate regression to answer *causal* research questions that follow the form "what is the effect of $X$ on $Y$?"

What is the effect of being in public institution ($X_i=1$) vs. private institution ($X_i=0$) on other off-campus expenses ($Y_i$) for students from the `df_ipeds_filtered` dataframe? 
    - $X$ var is `treatment`; $Y$ var is `read`

```{r}
df_ipeds_filtered %>% group_by(control) %>% summarize(
  mean_expenses = mean(oth_expense_off, na.rm = TRUE),
)
```
What is the effect of being at a private institution ($X$) on other off-campus expenses ($Y$) for full-time students?


```{r}
df_ipeds_filtered %>% filter(control == 1) %>% ggplot(aes(x=control, y=oth_expense_off)) + geom_point() + stat_smooth(method = 'lm')

summary(lm(formula = oth_expense_off ~ control, data = df_ipeds_filtered %>% filter(control==1)))

```

Why are we using expenses as dependent variables?

- Ordinary Least Squares (OLS) multivariate regression assumes continuous dependent variable; regression models with categorical dependent variables are a bit more difficult to interpret

## Bias and efficiency

Recall that the fundamental goal of inferential statistics is to make statements about **population parameters** by using **estimators** to calculate **point estimates** from sample data


- Desirable properties of our point estimates (e.g. $\hat{\beta}$ or $\bar{Y}$)
  - Efficient
  - Unbiased
  
### Efficiency (also called "precision")

- Definition: how close your point estimate(s) is to the population parameter
- Standard Error: on average, how far away is a point estimate from one random sample from the value of the population parameter
- _Therefore_, an efficient point estimate is one with a low standard error (in other words, the sampling distribution of $\beta_1$ has low variance or is "tight" around the population parameter)
  - often, the easiest way to increase efficiency (decrease standard errors) is to increase sample size
  
### Bias

- Bias: consistently overestimates or underestimates population parameter in repeated random samples
  - There are many different types of bias!
- _Sampling Bias_:
  - Estimate of population parameter is biased because you fail to take a random sample
  - Example: goal is to estimate high school graduation rate; take a random sample of 10th graders and see if they graduate within three years.
- _Omitted variable bias_:
  - Bias in estimate of $\beta_1$ due to omitting necessary "control" variables in your regression model


    











